First Order Alternation

### 2019-02-03 quantifiers/body.tex
ok

\begin{figure}[htb]
\vspace*{-\baselineskip}
\begin{center}
\begin{minipage}{5cm}
\input ta.pdf_t
\end{minipage} 
\begin{minipage}{5cm}
\footnotesize{\[A \left\{\begin{array}{rcl}
s_1(x,z) & \arrow{a(t)}{} & t \geq z \wedge s_2(z,t) \\
s_2(x,z) & \arrow{a(t)}{} & t \geq z \wedge z-x \geq 2 \wedge s_2(x,t)
\end{array}\right.\] 
\[B \left\{\begin{array}{rcl}
q_1(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge q_1(y,t) \\
q_1(y,u) & \arrow{a(t)}{} & t \geq u \wedge q_2(u,t) \\
q_2(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge u - y \neq 1 \wedge q_2(y,t)
\end{array}\right.\]
\[\overline{B}\left\{\begin{array}{rcl}
q_1(y,u) & \arrow{a(t),b(t)}{} & t < u \vee q_1(y,t) \\
q_1(y,u) & \arrow{a(t)}{} & t < u \vee q_2(u,t) \\
q_2(y,u) & \arrow{a(t),b(t)}{} & t < u \vee u - y = 1 \vee q_2(y,t)
\end{array}\right.\]}
\end{minipage}
\end{center}
\vspace*{-\baselineskip}
\caption{First Order Alternating Automata for Solving Inclusion of Timed Automata}
\label{fig:ta}
\vspace*{-\baselineskip}
\end{figure}

@@

The work probably closest to the one reported here concerns the model
of predicate automata (PA) \cite{Farzan15,Farzan16,KincaidPhD}, used
in the verification of parameterized concurrent programs with shared
memory. In this model, the alphabet consists of pairs of program
statements and thread identifiers, thus being infinite because the
number of threads is unbounded. Because thread identifiers can only be
compared for (dis-)equality, the data theory in PA is the theory of
equality. Even with this simplification, the emptiness problem is
undecidable when either the predicates have arity greater than one
\cite{Farzan15} or quantified transition rules
\cite{KincaidPhD}. Checking emptiness of quantifier free PA is
possible semi-algorithmically, by explicitly enumerating reachable
configurations and checking coverage by looking for permuations of
argument values. However, no semi-algorithm is given for quantified
PA.  Dealing with quantified transition rules is one of our
contributions.

@@

For two integers $0 \leq i \leq j$, we denote by $[i,j]$ the set
$\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[0,i]$. We consider two
sorts $\Data$ and $\Bool$, where $\Data$ is an infinite domain and
$\Bool = \set{\top,\bot}$ is the set of boolean values true ($\top$)
and false ($\bot$), respectively. The $\Data$ sort is equipped with
finitely many function symbols $f : \Data^{\#(f)} \rightarrow \Data$,
where $\#(f)\geq0$ denotes the number of arguments (arity) of
$f$. When $\#(f)=0$, we say that $f$ is a constant. A \emph{predicate}
is a function symbol $p : \Data^{\#(p)} \rightarrow \Bool$, denoting a
relation of arity $\#(p)$ and we write $\preds$ for the set of
predicates.

In the following, we shall consider that the interpretation of all
function symbols $f : \Data^{\#(f)} \rightarrow \Data$ that are not
predicates is fixed by the interpretation of the $\Data$ sort,
e.g.\ if $\Data$ is the set of integers $\zed$, the function symbols
are zero, the successor function and the arithmetic operations of
addition and multiplication. For simplicity, we further blur the
notational distinction between function symbols and their
interpretations.

Let $\vars = \set{x,y,z,\ldots}$ be an infinite countable set of
variables, ranging over $\Data$. Terms are either constants of sort
$\Data$, variables or function applications $f(t_1,\ldots,t_{\#(f)})$,
where $t_1,\ldots,t_{\#(f)}$ are terms. The set of first order
formulae is defined by the syntax below:
\[\phi := t \teq s \mid p(t_1,\ldots,t_{\#(p)}) 
\mid \neg \phi_1 \mid \phi_1 \wedge \phi_2 \mid \exists x ~.~
\phi_1 \] where $t,s,t_1,\ldots,t_{\#(p)}$ denote terms. We write
$\phi_1 \vee \phi_2$, $\phi_1 \rightarrow \phi_2$ and $\forall x ~.~
\phi_1$ for $\neg(\neg\phi_1 \wedge \neg\phi_2)$, $\neg\phi_1 \vee
\phi_2$ and $\neg\exists x ~.~ \neg\phi_1$, respectively. We denote by
$\fv{}{\phi}$ the set of free variables in $\phi$. The size
$\len{\phi}$ of a formula $\phi$ is the number of symbols needed to
write it down.

A \emph{sentence} is a formula $\phi$ in which each variable occurs
under the scope of a quantifier, i.e. $\fv{}{\phi} = \emptyset$.  A
formula is \emph{positive} if each predicate symbol occurs under an
even number of negations and we denote by $\posforms(Q,X)$ the set of
positive formulae with predicates from the set $Q \subseteq \preds$
and free variables from the set $X \subseteq \vars$.

A formula is in \emph{prenex form} if it is of the form $\varphi =
Q_1x_1 \ldots Q_nx_n ~.~ \phi$, where $\phi$ has no quantifiers. In
this case we call $\phi$ the \emph{matrix} of $\varphi$. Every
first order formula can be written in prenex form, by renaming each
quantified variable to a unique name and moving the quantifiers
upfront.

@@

An \emph{interpretation} $\I$ maps each predicate $p$ into a set $p^\I
\subseteq \Data^{\#(p)}$, if $\#(p)>0$, or into an element of $\Data$
if $\#(p)=0$. A \emph{valuation} $\nu$ maps each variable $x$ into an
element of $\Data$. Given a term $t$, we denote by $t^\nu$ the value
obtained by replacing each variable $x$ by the value $\nu(x)$ and
evaluating each function application. For a formula $\phi$, we define
the forcing relation $\I,\nu \models \phi$ recursively on the
structure of $\phi$, as usual. 
\ifLongVersion
\[
\begin{array}{rcl}
\I,\nu \models t \teq s & \iff & t^\nu=s^\nu \\
\I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I \\
\I,\nu \models \neg\phi_1 & \iff & \I,\nu \not\models \phi_1 \\
\I,\nu \models \phi_1 \wedge \phi_2 & \iff & \I,\nu \models \phi_i \text{, for all } i=1,2 \\ 
\I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data 
\end{array}
\]
where $\nu[x\leftarrow d]$ is the valuation which assigns $d$ to $x$
and behaves like $\nu$ elsewhere.  
\fi 
For a formula $\phi$ and a valuation $\nu$, we define $\sem{\phi}_\nu
\isdef \set{\I \mid \I,\nu \models \phi}$ and drop the $\nu$ subscript
for sentences. A sentence $\phi$ is \emph{satisfiable}
(\emph{unsatisfiable}) if $\sem{\phi}\neq\emptyset$
($\sem{\phi}=\emptyset$). An element of $\sem{\phi}$ is called a
\emph{model} of $\phi$. A formula $\phi$ is \emph{valid} if $\I,\nu
\models \phi$ for every interpretation $\I$ and every valuation
$\nu$. For two formulae $\phi$ and $\psi$ we write $\phi \models \psi$
for $\sem{\phi} \subseteq \sem{\psi}$, in which case we say that
$\phi$ \emph{entails} $\psi$.

Interpretations are partially ordered by the pointwise subset order,
defined as $\I_1 \subseteq \I_2$ if and only if $p^{\I_1} \subseteq
p^{\I_2}$ for each predicate $p \in \preds$. Given a set $\mathcal{S}$
of interpretations, a minimal element $\I \in \mathcal{S}$ is an
interpretation such that for no other interpretation $\I' \in
\mathcal{S} \setminus \set{\I}$ do we have $\I' \subseteq \I$. For a
formula $\phi$ and a valuation $\nu$, we denote by $\minsem{\phi}_\nu$
and $\minsem{\phi}$ the set of minimal interpretations from
$\sem{\phi}_\nu$ and $\sem{\phi}$, respectively.

@@

Let $\Sigma$ be a finite alphabet $\Sigma$ of \emph{input
  events}. Given a finite set of variables $X \subseteq \vars$, we
denote by $X \mapsto \Data$ the set of valuations of the variables $X$
and $\Sigma[X] = \Sigma \times (X \mapsto \Data)$ be the possibly
infinite set of \emph{data symbols} $(a,\nu)$, where $a$ is an input
symbol and $\nu$ is a valuation. A \emph{data word} (simply called
word in the following) is a finite sequence $w=(a_1,\nu_1)(a_2,\nu_2)
\ldots (a_n,\nu_n)$ of data symbols. Given a word $w$, we denote by
$\event{w} \isdef a_1\ldots a_n$ its sequence of input events and by
$\data{w}$ the valuation associating each time-stamped variable
$\stamp{x}{i}$ the value $\nu_i(x)$, for all $x \in \vars$ and
$i\in[1,n]$. We denote by $\varepsilon$ the empty sequence, by
$\Sigma^*$ the set of finite sequences of input events and by
$\Sigma[X]^*$ the set of data words over the variables $X$.

Formally, a \emph{first order alternating automaton} is a tuple $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is a finite set of
input events, $X$ is a finite set of input variables, $Q$ is a finite
set of predicates denoting control states, $\iota \in
\posforms(Q,\emptyset)$ is a sentence defining initial configurations,
$F \subseteq Q$ is the set of predicates denoting final states, and
$\Delta$ is a set of \emph{transition rules} of the form
\(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is a
predicate, $a \in \Sigma$ is an input event and $\psi \in
\posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive formula,
where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$. The quantifiers
occurring in the right-hand side formula of a transition rule are
referred to as \emph{transition quantifiers}. The \emph{size} of $\A$
is defined as $\len{\A} = \len{\iota} +
\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}}
\len{\psi}$.

@@

The execution semantics of automata is given in close analogy with the
case of boolean alternating automata, with transition rules of the
form $q \arrow{a}{} \phi$, where $q$ is a boolean constant and $\phi$
a positive boolean combination of such constants. For instance, $q_0
\arrow{a}{} q_1 \wedge q_2 \vee q_3$ means that the automaton can
choose to transition in either both $q_1$ and $q_2$ or in $q_3$
alone. This intuition leads to saying that the steps of the automaton
are defined by the minimal boolean models of the transition
formulae. In this case, both $\set{q_1 \leftarrow \top, q_2 \leftarrow
  \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow \bot, q_2
  \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models, however
$\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$,
is a model but is not minimal. The original definition of alternating
finite-state automata \cite{ChandraKozenStockmeyer81} works around
this problem by considering boolean valuations (models) instead of
formulae. However, describing first-order alternating automata using
interpretations instead of formulae would be rather hard to follow.

Given a predicate $q \in Q$ and a tuple of data values
$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
a \emph{configuration}\footnote{Note that a configuration is not a
  logical term since data values cannot be written in logic.}. To
formalize the execution semantics of automata, we relate sets of
configurations to models of first order sentences, as follows. Each
first-order interpretation $\I$ corresponds to a set of configurations
$\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid q \in Q,~
  \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a \emph{cube}. For a
set $\mathcal{S}$ of interpretations, we define $\cube{\mathcal{S}}
\isdef \set{\cube{\I} \mid \I \in \mathcal{S}}$.

\begin{definition}\label{def:execution}
Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$ and a cube $c$, an
\emph{execution} of $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$, starting with $c$, is
a (possibly infinite) forest $\T = \set{T_1,T_2,\ldots}$, where each
$T_i$ is a tree labeled with configurations, such that:
\begin{compactenum}
\item\label{it1:execution} $c = \set{T(\epsilon) \mid T \in \T}$ is the set of
  configurations labeling the roots of $T_1,T_2,\ldots$ and
%
\item\label{it2:execution} if $q(d_1,\ldots,d_{\#(q)})$ labels a node on the level $j \in
  [n-1]$ in $T_i$, then the labels of its children form a cube from
  $\cube{\minsem{\psi}_{\eta}}$, where $\eta = \nu_{j+1}[y_1
    \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$ and
  \(q(y_1,\ldots,y_{\#(q)}) \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is
  a transition rule of $\A$.
\end{compactenum}
\end{definition}

\begin{definition}\label{def:accepting}
An execution $\T$ over $w$, starting with $c$, is \emph{accepting} if
and only if \begin{compactitem}
\item\label{it1:accepting} all paths in $\T$ have the same length $n$,
  and
  %
\item\label{it2:accepting} the frontier of each tree $T \in \T$ is
  labeled with \emph{final configurations} $q(d_1,\ldots,d_{\#(q)})$,
  where $q \in F$.
\end{compactitem} 
If $\A$ has an accepting execution over $w$ starting with a cube $c
\in \cube{\minsem{\iota}}$, then $\A$ \emph{accepts} $w$ and let
$\lang{\A}$ be the set of words accepted by $\A$.
\end{definition}
In this paper, we address the following questions: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$?
\end{compactenum}

\subsection{Symbolic Execution}

In the upcoming developments it is sometimes more convenient to work
with logical formulae defining executions of automata, than with
low-level execution forests. For this reason, we first introduce
\emph{path formulae} $\pathform{\alpha}$, which are formulae defining
the executions of an automaton, over words that share a given sequence
$\alpha$ of input events.  Second, we restrict a path formula
$\pathform{\alpha}$ to an \emph{acceptance formula}
$\accform{\alpha}$, which defines only accepting executions over words
that share a given input sequence. Otherwise stated,
$\accform{\alpha}$ is satisfiable if and only if the automaton accepts
a word $w$ such that $\event{w} = \alpha$.

Let $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the
rest of this section. For any $i \in \nat$, we denote by $\stamp{Q}{i}
= \set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} =
\set{\stamp{x}{i} \mid x \in X}$ the sets of time-stamped predicates
and variables, respectively. As a shorthand, we write $\stamp{Q}{\leq
  n}$ (resp.  $\stamp{X}{\leq n}$) for the set $\set{\stamp{q}{i} \mid
  q \in Q, i \in [n]}$ (resp. $\set{\stamp{x}{i} \mid x \in X, i \in
  [n]}$). For a formula $\psi$ and $i \in \nat$, we define
$\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the
formula in which all input variables and state predicates (and only
those symbols) are replaced by their time-stamped counterparts. As a
shorthand, we shall write $q(\vec{y})$ for $q(y_1,\ldots,y_{\#(q)})$,
when no confusion arises.

>>>>

\begin{figure}[htb]
\vspace*{-\baselineskip}
\begin{center}
\begin{minipage}{5cm}
\input ta.pdf_t
\end{minipage} 
\begin{minipage}{5cm}
\footnotesize{\[A \left\{\begin{array}{rcl}
s_1(x,z) & \arrow{a(t)}{} & t \geq z \wedge s_2(z,t) \\
s_2(x,z) & \arrow{a(t)}{} & t \geq z \wedge z-x \geq 2 \wedge s_2(x,t)
\end{array}\right.\] 
\[B \left\{\begin{array}{rcl}
q_1(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge q_1(y,t) \\
q_1(y,u) & \arrow{a(t)}{} & t \geq u \wedge q_2(u,t) \\
q_2(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge u - y \neq 1 \wedge q_2(y,t)
\end{array}\right.\]
\[\overline{B}\left\{\begin{array}{rcl}
q_1(y,u) & \arrow{a(t),b(t)}{} & t < u \vee q_1(y,t) \\
q_1(y,u) & \arrow{a(t)}{} & t < u \vee q_2(u,t) \\
q_2(y,u) & \arrow{a(t),b(t)}{} & t < u \vee u - y = 1 \vee q_2(y,t)
\end{array}\right.\]}
\end{minipage}
\end{center}
\vspace*{-\baselineskip}
\caption{First Order Alternating Automata Checking Inclusion of Timed Automata}
\label{fig:ta}
\vspace*{-\baselineskip}
\end{figure}

@@

The work probably closest to ours concerns the model of predicate
automata (PA) \cite{Farzan15,Farzan16,KincaidPhD}, used in the
verification of parameterized concurrent programs with shared
memory. In this model, the alphabet consists of pairs of program
statements and thread identifiers, thus being infinite because the
number of threads is unbounded. Because thread identifiers can only be
compared for (dis-)equality, the data theory in PA is the theory of
equality. Even with this simplification, the emptiness problem is
undecidable when either the predicates have arity greater than one
\cite{Farzan15} or quantified transition rules
\cite{KincaidPhD}. Checking emptiness of quantifier free PA is
possible semi-algorithmically, by explicitly enumerating reachable
configurations and checking coverage by looking for permuations of
argument values. However, no semi-algorithm is given for quantified
PA.  Dealing with quantified transition rules is one of our
contributions.

@@

For two integers $0 \leq i \leq j$, we define $[i,j] \isdef
\set{i,\ldots,j}$ and $[i] \isdef [0,i]$. We consider two disjoint
sorts $\Data$ and $\Bool$, where $\Data$ is an infinite domain and
$\Bool = \set{\top,\bot}$ is the set of boolean values true ($\top$)
and false ($\bot$), respectively. The $\Data$ sort is equipped with
countably many function symbols $f : \Data^{\#(f)} \rightarrow \Data
\cup \Bool$, where $\#(f)\geq0$ denotes the number of arguments
(arity) of $f$. A \emph{predicate} is a function symbol $p :
\Data^{\#(p)} \rightarrow \Bool$ that is, a $\#(p)$-ary relation.

We consider the interpretation of all function symbols $f :
\Data^{\#(f)} \rightarrow \Data$ to be fixed by the interpretation of
the $\Data$ sort, for instance if $\Data$ is the set of integers
$\zed$, these are zero, the successor function and the arithmetic
operations of addition and multiplication. We extend this convention
to several predicates over $\Data$, such as the inequality relation
over $\zed$, and write $\preds$ for the set of \emph{uninterpreted
  predicates}.

Let $\vars = \set{x,y,z,\ldots}$ be a countably infinite set of
variables, ranging over $\Data$. Terms are either constants of sort
$\Data$, variables or function applications $f(t_1,\ldots,t_{\#(f)})$,
where $t_1,\ldots,t_{\#(f)}$ are terms. The set of first order
formulae is defined by the syntax below:
\[\phi := t = s \mid p(t_1,\ldots,t_{\#(p)}) 
\mid \neg \phi_1 \mid \phi_1 \wedge \phi_2 \mid \exists x ~.~
\phi_1 \] where $t,s,t_1,\ldots,t_{\#(p)}$ denote terms and $p$ is a
predicate symbol. We write $\phi_1 \vee \phi_2$, $\phi_1 \rightarrow
\phi_2$ and $\forall x ~.~ \phi_1$ for $\neg(\neg\phi_1 \wedge
\neg\phi_2)$, $\neg\phi_1 \vee \phi_2$ and $\neg\exists x ~.~
\neg\phi_1$, respectively. $\fv{}{\phi}$ is the set of free variables
in $\phi$ and the size $\len{\phi}$ of a formula $\phi$ is the number
of symbols needed to write it down. A \emph{sentence} is a formula
$\phi$ with no free variables. A formula is \emph{positive} if each
uninterpreted predicate symbol occurs under an even number of
negations and we denote by $\posforms(Q,X)$ the set of positive
formulae with predicates from the set $Q \subseteq \preds$ and free
variables from the set $X \subseteq \vars$. A formula is in
\emph{prenex form} if it is of the form $\varphi = Q_1x_1 \ldots
Q_nx_n ~.~ \phi$, where $\phi$ has no quantifiers. In this case we
call $\phi$ the \emph{matrix} of $\varphi$. Every first order formula
can be written in prenex form, by renaming each quantified variable to
a unique name and moving the quantifiers upfront.

@@

An \emph{interpretation} $\I$ maps each predicate symbol $p$ into a
set $p^\I \subseteq \Data^{\#(p)}$, if $\#(p)>0$, or into an element
of $\Bool$ if $\#(p)=0$. A \emph{valuation} $\nu$ maps each variable
$x$ into an element of $\Data$. Given a term $t$, we denote by $t^\nu$
the value obtained by replacing each variable $x$ by the value
$\nu(x)$ and evaluating each function application. For a formula
$\phi$, we define the forcing relation $\I,\nu \models \phi$
recursively on the structure of $\phi$, as usual.  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\ifLongVersion
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\[
\begin{array}{rcl}
\I,\nu \models t \teq s & \iff & t^\nu=s^\nu \\
\I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I \\
\I,\nu \models \neg\phi_1 & \iff & \I,\nu \not\models \phi_1 \\
\I,\nu \models \phi_1 \wedge \phi_2 & \iff & \I,\nu \models \phi_i \text{, for all } i=1,2 \\ 
\I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data 
\end{array}
\]
where $\nu[x\leftarrow d]$ is the valuation which assigns $d$ to $x$
and behaves like $\nu$ elsewhere.  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\fi 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For a formula $\phi$ and a valuation $\nu$, we define $\sem{\phi}_\nu
\isdef \set{\I \mid \I,\nu \models \phi}$ and drop the $\nu$ subscript
for sentences. A sentence $\phi$ is \emph{satisfiable} if
$\sem{\phi}\neq\emptyset$. An element of $\sem{\phi}$ is called a
\emph{model} of $\phi$. A formula $\phi$ is \emph{valid} if $\I,\nu
\models \phi$ for every interpretation $\I$ and every valuation
$\nu$. We say that $\phi$ \emph{entails} $\psi$, written $\phi \models
\psi$ if and only if $\sem{\phi} \subseteq \sem{\psi}$.

Interpretations are partially ordered by the pointwise subset order,
defined as $\I_1 \subseteq \I_2$ if and only if $p^{\I_1} \subseteq
p^{\I_2}$ for each predicate symbol $p \in \preds$. Given a formula
$\phi$ and a valuation $\nu$, we define $\minsem{\phi}_\nu \isdef
\set{\I \mid \I,\nu \models \phi,~ \forall \I' \subseteq \I ~.~
  \I',\nu \not\models \phi}$ the set of minimal interpretations that,
together with $\nu$, form models of $\phi$.

@@

Let $\Sigma$ be a finite alphabet $\Sigma$ of \emph{input
  events}. Given a finite set of variables $X \subseteq \vars$, we
denote by $X \mapsto \Data$ the set of valuations of the variables $X$
and $\Sigma[X] = \Sigma \times (X \mapsto \Data)$ be the possibly
infinite set of \emph{data symbols} $(a,\nu)$, where $a$ is an input
symbol and $\nu$ is a valuation. A \emph{data word} (simply called
word in the following) is a finite sequence $w=(a_1,\nu_1)(a_2,\nu_2)
\ldots (a_n,\nu_n)$ of data symbols. Given a word $w$, we denote by
$\event{w} \isdef a_1\ldots a_n$ its sequence of input events and by
$\data{w}$ the valuation associating each time-stamped variable
$\stamp{x}{i}$, where $x \in \vars$, the value $\nu_i(x)$, for all
$i\in[1,n]$. We denote by $\varepsilon$ the empty sequence, by
$\Sigma^*$ the set of finite input sequences and by $\Sigma[X]^*$ the
set of finite data words over the variables $X$.

A \emph{first order alternating automaton} is a tuple $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is a finite set of
input events, $X$ is a finite set of input variables, $Q$ is a finite
set of predicates denoting control states, $\iota \in
\posforms(Q,\emptyset)$ is a sentence defining initial configurations,
$F \subseteq Q$ is the set of predicates denoting final states, and
$\Delta$ is a set of \emph{transition rules} of the form
\(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is a
predicate, $a \in \Sigma$ is an input event and $\psi \in
\posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive formula,
where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$. The quantifiers
occurring in the right-hand side formula of a transition rule are
referred to as \emph{transition quantifiers}. The \emph{size} of $\A$
is $\len{\A} \isdef \len{\iota} + \sum \set{\len{\psi} \mid q(\vec{y})
  \arrow{a(X)}{} \psi \in \Delta}$.

@@

The execution semantics of automata is given in close analogy with the
classical definition of propositional alternating automata, with
transition rules of the form $q \arrow{a}{} \phi$, where $q$ is a
propositional variable and $\phi$ a positive boolean combination of
propositional variables. For instance, $q_0 \arrow{a}{} q_1 \wedge q_2
\vee q_3$ means that the automaton can choose to transition in either
both $q_1$ and $q_2$ or in $q_3$ alone. This intuition leads to saying
that a transition is a \emph{minimal model} of a right hand side
formula of a rule. In our example, both $\set{q_1 \leftarrow \top, q_2
  \leftarrow \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow
  \bot, q_2 \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models,
however $\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow
  \top}$, is a model but is not minimal. The original definition of
alternating finite-state automata \cite{ChandraKozenStockmeyer81}
works around this problem by considering boolean valuations (models)
instead of formulae. However, a finite description of a first order
alternating automaton must use formulae, because there can be
infinitely many satisfying interpretations that are incomparable
w.r.t. the pointwise subset ordering.

Given an uninterpreted predicate symbol $q \in Q$ and a tuple of data
values $d_1,\ldots,d_{\#(q)}$, the tuple $(q, d_1,\ldots,d_{\#(q)})$
is called a \emph{configuration}, sometimes written
$q(d_1,\ldots,d_{\#(q)})$, when no confusion arises. A configuration
is \emph{final} if $q \in F$. An interpretation $\I$ corresponds to a
set of configurations $\cube{\I} \isdef \set{(q, d_1,\ldots,d_{\#(q)})
  \mid q \in Q,~ (d_1,\ldots,d_{\#(q)}) \in q^\I}$, called a
\emph{cube}. This notation is lifted to sets of configurations in the
usual way.

\begin{definition}\label{def:execution}
Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$ and a
cube $c$, an \emph{execution} of
$\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$, starting with $c$, is
a forest $\T = \set{T_1,T_2,\ldots}$, where each $T_i$ is a tree
labeled with configurations, such that:
\begin{compactenum}
\item\label{it1:execution} $c = \set{T(\epsilon) \mid T \in \T}$ is the set of
  configurations labeling the roots of $T_1,T_2,\ldots$ and
%
\item\label{it2:execution} if $(q,d_1,\ldots,d_{\#(q)})$ labels a node
  on the level $j \in [n-1]$ in $T_i$, then the labels of its children
  form a cube from $\cube{\minsem{\psi}_{\eta}}$, where $\eta =
  \nu_{j+1}[y_1 \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$
  and \(q(y_1,\ldots,y_{\#(q)}) \arrow{a_{j+1}(X)}{} \psi \in \Delta\)
  is a transition rule of $\A$.
\end{compactenum}
\end{definition}
An execution $\T$ over $w$, starting with $c$, is \emph{accepting} if
and only if all paths in $\T$ have the same length and the frontier of
each tree $T \in \T$ is labeled with final configurations.  If $\A$
has an accepting execution over $w$ starting with a cube $c \in
\cube{\minsem{\iota}}$, then $\A$ \emph{accepts} $w$ and let
$\lang{\A}$ be the set of words accepted by $\A$. We consider the
following problems: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$ ?
\end{compactenum}
For technical reasons, we address the following \emph{relative
  acceptance} problem first: given an automaton $\A$ and an input
sequence $\alpha \in \Sigma^*$, does there exists a word $w \in
\lang{\A}$ such that $\event{w} = \alpha$ ?  By solving this problem
first, we develop the machinery required to prove that first order
alternating automata are closed under all boolean operations and,
further, set up the ground for a semi-algorithm for the emptiness
problem.

\subsection{Path Formulae}

In the upcoming developments it is sometimes more convenient to work
with logical formulae defining executions of automata, than with
low-level execution forests. For this reason, we first introduce
\emph{path formulae} $\pathform{\alpha}$, which are formulae defining
the executions of an automaton, over words that share a given sequence
$\alpha$ of input events.  Second, we restrict a path formula
$\pathform{\alpha}$ to an \emph{acceptance formula}
$\accform{\alpha}$, which defines only accepting executions over words
that share a given input sequence. Otherwise stated,
$\accform{\alpha}$ is satisfiable if and only if the automaton accepts
a word $w$ such that $\event{w} = \alpha$. We reduce thus the relative
acceptance problem to the satisfiability of an effectively computable
formula in the first order theory of data with uninterpreted predicate
symbols.

Let $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the
rest of this section. For any $i \in \nat$, we denote by $\stamp{Q}{i}
= \set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} =
\set{\stamp{x}{i} \mid x \in X}$ the sets of time-stamped predicate
symbols and variables, respectively. As a shorthand, we write
$\stamp{Q}{\leq n}$ (resp.  $\stamp{X}{\leq n}$) for the set
$\set{\stamp{q}{i} \mid q \in Q, i \in [n]}$ (resp. $\set{\stamp{x}{i}
  \mid x \in X, i \in [n]}$). For a formula $\psi$ and $i \in \nat$,
we define $\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$
the formula in which all input variables and state predicates (and
only those symbols) are replaced by their time-stamped
counterparts. As a shorthand, we shall write $q(\vec{y})$ for
$q(y_1,\ldots,y_{\#(q)})$, when no confusion arises.

Given a sequence of input events $\alpha = a_1 \ldots a_n \in
\Sigma^*$, the \emph{path formula} of $\alpha$ is:
\begin{equation}\label{eq:pathform}
\begin{array}{c}
\pathform{\alpha} \isdef \stamp{\iota}{0} \wedge 
\bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{}
  \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
\stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i}
\end{array}
\end{equation}
The automaton $\A$, to which $\pathform{\alpha}$ refers, will always
be clear from the context. To formalize the relation between the
low-level configuration-based execution semantics and path formulae,
consider a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in
\Sigma[X]^*$. Any execution $\T$ of $\A$ over $w$ has an associated
interpretation $\I_{\T}$ of time-stamped predicates $\stamp{Q}{\leq
  n}$:
\[\I_{\T}(\stamp{q}{i}) \isdef \set{(d_1,\ldots,d_{\#(q)}) \mid
  (q,d_1,\ldots,d_{\#(q)}) \text{ labels a node on level $i$ in $\T$}}
,~\forall q\in Q ~\forall i \in [n]\]
